A Formal Logic for Formal Category Theory

### 2022-12-08 paper/reviews/response.txt

We thank the reviewers for their very detailed and helpful feedback.  We
will integrate them into the final version of the paper --- including
all of the clarification comments not mentioned below.

We respond to the most pertinent questions first. And each response
includes "Reviewer X" so you can search for the responses to your
specific questions.

We agree with Review 3 that the most important advantages of our
approach are for enriched category theory (indeed, this was our
motivation for starting this work), and can revise the introduction to
emphasize this aspect more.  However, we do feel there are some
advantages even for ordinary category theory: while a synthetic theory
cannot eliminate all side conditions, it can help structure when the
user really needs to prove them.  One of the examples the reviewer
mentions, respect for quotients, is a classic example where an
"analytic" approach is unpleasant in practice ("setoid hell"), and
improved by synthetic approaches in NuPRL/Observational Type
Theory/HoTT/Cubical type theory.  While an adept category theorist may
know how to manage these obligations and e.g. pick a definition that
is more automatically natural, our experience explaining category
theory to programming languages researchers as background for
understanding some of our past papers suggests that the support of a
type theory may be useful for explaining how experts think about these
issues. We agree with the reviewer that the form the "domain specific
proof assistant" could take is as a library design for an existing
proof assistant, i.e., an embedded DSL.

Review 3 asks what the relevant CwF structures are in Constructions 4
and 5.  Our type theory distinguishes between the virtual equipment
layer (categories, functors, profunctors, natural transformations) and
the "meta" layer, a dependent type theory.  The semantics interprets
this in a hyperdoctrine of virtual equipments, with the CwF structure
for the meta-layer on the base of the hyperdoctrine.  In the "standard"
model, the base is just the category of sets Set, while the fibers of
the hyperdoctrine are Cat.  So there is no need for a CwF structure on
Cat.

Review 3 asks if every CwF gives rise to a HVE. Certainly construction
4 can be adapted to any elementary topos with sufficient universes
since the category-theoretic constructions needed are all
intuitionistically valid. So this should correspond to a Category with
Families extended with certain universal properties, for instance,
extensional identity types.

Review 3 suggests some additional references and examples (Palmgren,
  Weber, Di Liberti and Loregian). Thank you; we will include them!

Review 1 asks about the limitations of the present formalism, and
Review 2 similarly asked about why we restrict to unary
functors/binary profunctors. There are several useful concepts we
cannot formalize because our judgments are limited to covariant
functors of a single argument: categories of functors and opposite
categories, for instance. Next steps for the current work would be to
extend the judgmental structure to support these. Some preliminary
work along these lines is in Reference [21]. However, there is a
tradeoff: by adding more structure in the type theory, we support
fewer models. We chose these restrictions as a starting point because
they correspond to a well-understood notion of model (virtual
equipments), future extensions would likely require the equipment to
come with a monoidal closed structure. In particular, we expect that
categories enriched in a non-symmetric monoidal closed category will
not be a model of these extensions.

Review 1 asks about the definition of d(Ξ) and the presheaf
elimination rules. d-(x:C) and d+(x:C) are both x:C; we will add this
definition. The negative presheaf elimination rule is the closure
under substitution of a : C ; p : P- C |- a in p Set i.e. evaluation
is a map C^op x (Set^{C^op}) -> Set

Review 1 asks about the tensor product of profunctors. The tensor
product of profunctors is indeed a colimit --- e.g. it can be defined
using a coend.  Syntactically, beta is just a bound variable (and can
of course be alpha-converted).  The fact that maps out of the tensor
respect the quotienting is implicit/synthetic, and comes from the fact
that all open terms ... beta : C ... |- t : S respect this
quotienting.  This is similar to how quotient types are handled in
NuPRL/Observational type theory/HoTT/Cubical type theory.

Review 1 had several questions about Section 4. We will rephrase
Theorem 2 with less shorthand. The formal meaning of "almost any" is
to emphasize that the initiality theorem is modular in that most
connectives are independent of each other. The exception is that to
get the right universal property for hom and tensor of profunctors, we
need the exponential of profunctors. vDbl_r and Cat_U should, as the
reviewer guessed, be vEq and Cat_κ. The purpose of Construction 3 is
to reduce the construction of a model of VETT to just constructing a
virtual equipment, since any virtual equipment can be "upgraded" to a
hyperdoctrine of equipments using this construction.

Review 2 asks if one can generate explicit proofs of functoriality
/naturality from proofs in the type theory.  Yes, "running" the
(constructive) proof of soundness (Theorem 2) would do exactly this, by
compiling our "synthetic" proofs to "analytic" ones.

Review 2 asks about "P ▹∀α Q" and the difference with "∀α.P".

P ▹∀α Q is the 'hom' of profunctors, the right adjoint to - (x)∃α P.  This
type binds both an object variable α and a transformation variable x:P.

In the type of composition 
   ∀α1. (α1 → α2) |>∀α2 (α2 → α3 |>∀α3 (α1 → α3))
α2 occurs covariantly in (α1 → α2)
and covariantly in (α2 → α3 |>∀α3 (α1 → α3)) because the
contravariance of |> and → cancel out.

∀α.P is the 'end' over a profunctor with 1 free variable (Figure 5,
bottom row).  This type is used to bind the "last" object variable in a
singleton context. The type for functoriality that the reviewer gives is
isomorphic to the type we gave by Lemma 3, part 5.

Review 2 asks about our comments on the Yoneda lemma. We had a typo in
this formula, thanks --- we meant ∀x. [ (∀y.(x=y) ⇒ Py) iff P x].

Review 2 asks about the proof of initiality theorem. The initiality
theory could in principle be mechanized in Agda, though we have not
yet done so.

### 2023-01-27 paper/paper.tex
final version?

\begin{document}
%% \newtheorem{theorem}{Theorem}
%% \newtheorem{lemma}{Lemma}
\newtheorem{construction}{Construction}
%% \newtheorem{definition}{Definition}
\newcommand{\vett}{VETT}
\newcommand{\ohol}{\vett}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\pto}{\nrightarrow}
\newcommand{\pfrom}{\nleftarrow}
\newcommand{\vcat}{\mathcal}
\newcommand{\cat}{\mathbbm}
\newcommand{\isaSet}{\,\,\textrm{Set}}
\newcommand{\isaTy}{\,\,\textrm{Type}}
\newcommand{\isaCat}{\,\,\textrm{Cat}}
\newcommand{\isSmall}{\,\,\textrm{Small}}
\newcommand{\Set}{\textrm{Set}}
\newcommand{\MonCat}{\textrm{MonCat}}
\newcommand{\SymMonCat}{\textrm{SymMonCat}}
\newcommand{\Syn}{\textrm{Syn}}
\newcommand{\vtkmnd}{\mathbb{K}\text{Mod} (\vcat{V},T)}
\newcommand{\rmod}{\text{RMod}}
\newcommand{\lmod}{\text{LMod}}

@@

\newcommand{\odotexists}[1]{\mathop{\overset{\exists #1}\odot}}
\newcommand{\tensorexistsXwithYandZ}[3]{#2 \odotexists{#1} #3}
\newcommand{\tensorintroofXandY}[2]{\textrm{pair}_\odot(#1,#2)}
\newcommand{\tensorintroatXwithYandZ}[3]{(#2,#1,#3)}
\newcommand{\tensorelimXtoYwithkontZ}[3]{\textrm{ind}_{\to}(#1,#2,#3)}
\newcommand{\tensorelimfromWXtoYwithkontZ}[4]{\textrm{ind}_{\odot}(#1,#2,#3,#4)}
\newcommand{\tensorelimVatWtoXtoYwithkontZ}[5]{\textrm{ind}_{\odot}(#4,#2,#1,#3,#5)}
\newcommand{\tensorelimWfromXtoYkontZ}[4]{\textrm{ind}_{\odot}(#4;#1,#2,#3)}
\newcommand{\tensorelimWkontZ}[2]{\textrm{ind}_{\odot}(#2;#1)}

@@

\title{A Formal Logic for Formal Category Theory}
\author{Max S. New\inst{1,2} and Daniel R. Licata\inst{2}}
\institute{University of Michigan \and Wesleyan University}

@@

\begin{abstract}
  We present a domain-specific type theory for constructions and proofs
  in category theory. The type theory axiomatizes notions of category,
  functor, profunctor and a generalized form of natural
  transformations. The type theory imposes an ordered linear restriction
  on standard predicate logic, which guarantees that all functions
  between categories are functorial, all relations are profunctorial,
  and all transformations are natural by construction, with no separate
  proofs necessary. Important category theoretic proofs such as the
  Yoneda lemma and Co-yoneda lemma become simple type theoretic proofs
  about the relationship between unit, tensor and (ordered) function
  types, and can be seen to be ordered refinements of theorems in
  predicate logic.  The type theory is sound and complete for a
  categorical model in \emph{virtual equipments}, which model both
  internal and enriched category theory. While the proofs in our type
  theory look like standard set-based arguments, the syntactic
  discipline ensure that all proofs and constructions carry over to
  enriched and internal settings as well.
\end{abstract}

@@

Category theory is a branch of mathematics that studies
higher-dimensional typed algebraic structures.
%
Originally developed for applications to homological algebra, it was
quickly discovered that categorical structures were common in logic and
computer science.  Formal systems like logics, type theories and
programming languages typically have sound and complete models given by
notions of structured categories~\cite{Lawvere69,LambekScott,moggi}.
%
This Curry-Howard-Lambek correspondence~
applies to simply typed lambda calculus~\cite{LambekScott},
computational lambda calculus~\cite{moggi}, linear logic~\cite{girard}
dependent type theory~\cite{cartmell1986}, and many other type theories
designed based on category-theoretic semantics.
%
The syntax of a type theory should present an initial object in its
category of models, a category-theoretic reformulation of logical
soundness and completeness.

While this research program has been quite successful,
category-theoretic techniques in computer science are sometimes
criticized for being overly complex and hard to learn, especially for
computer scientists who do not have a strong background in pure
mathematics.
%
While in some ways this could be remedied by better introductory
materials, in some ways category theory is objectively complex.
%
In a traditional set-theoretic formulation, notions such as adjoint
functors and limits produce a proliferation of ``naturality'' and
``functoriality'' side-conditions that must be discharged.
%
For example, when constructing an adjoint pair of
functors between two categories, a na\"ive approach would define all of
the data of the action on objects, action on arrows, prove the
functoriality of such actions, as well as construct two families of
transformations, prove they are natural and then finally proving a pair
of equalities relating compositions of natural transformations.
%
Carrying out these proofs explicitly is quite tedious and many newcomers
to are left with the impression that category theory is full of long,
but ultimately trivial constructions.
%
This complexity is compounded when we move from ordinary category theory to
enriched and internal category theory, where constructions must be
additionally proven continuous, monotone, etc, in addition to natural
or functorial.
%
However, these generalizations are often exactly what is needed for programming
language applications; for example, domain- and metric- and
step-index-enriched categories have been used to model recursive
programming languages and internal categories have been used to model
parametricity and gradual
typing~\cite{wand197913,topos-of-trees,reflexive-graphs,double-cats-gradual-typing}.

Fortunately, the tools of category theory itself can be employed to
simplify this complexity, specifically the tools of \emph{higher}
category theory.
%
As an analogy in differential calculus, when an adept analyst writes
down a function, they do not expand out the $\epsilon\mathord{-}\delta$
definition of continuity for a function and proceed from first
principles, but rather use certain \emph{syntactic principles} for
defining functions that are continuous by construction --- e.g. that
composition of continuous functions is continuous and that previously
defined functions are known to be continuous.
%
Similar principles apply to category theory itself: functors and
natural transformations are closed under composition and whiskering
operations, and experienced category theorists rely on these syntactic
principles to eliminate the tedium of explicit proofs.
%
In the case of category theory, these principles can be formalized
using algebraic structures such as 2-categories, bicategories,
(virtual) double categories and pro-arrow equipments~\cite{bicategories,proarrow-equipments,LEINSTER2002391,Cruttwell2010}, an
approach known as \emph{formal category theory}.
%
In these structures, rather than defining notions of category, functor
and natural transformation from first principles, they are axiomatized
in a manner similar to how a category axiomatizes a notion of space
and homomorphism.
%
Proofs in formal category theory apply to enriched and internal
settings, which are instances of the formal axioms.
%
A downside is that these algebraic structures are quite complicated, and
practitioners typically employ either an algebraic combinator syntax or
a 2-dimensional diagrammatic language that can be quite beautiful and
elegant, but is also somewhat removed from the traditional formulation
of category theory in terms of sets and functions.

@@

%% The type theoretic syntax allows us to write all of our 
%% constructions are written as if they were
%% ordinary functors and profunctors in set-based category theory.

@@

The syntax of \vett{} is an indexed, ordered linear, proof-relevant
variant of predicate logic over a unary type theory.
%
Just as a predicate logic has a notion of type, term, relation and
implication, \vett{} is based on four analogous category theoretic
concepts: categories, functors, profunctors and natural transformations
of profunctors.
%
Categories are treated like types, and the unary functors we consider in
this paper are each represented by a term whose type is a category and
whose one free variable ranges over a category.
%
The analogue of a relation is a \emph{profunctor} (defined below), which
is written like a set with free category variables.  Like the
restriction to unary functors, we restrict to profunctors with two free
variables.
%
The logic is proof-relevant in that the implications of relations are
generalized to natural transformations of profunctors, and we use a
$\lambda$-calculus notation to describe these ``proof terms''.
%
This analogy to predicate logic can be made formal: any construction in
\vett{} can be erased to a corresponding construction or proof in
predicate logic, as sets, functions, relations, and implication of
relations define a (somewhat degenerate) virtual equipment.
%% %
%% Our syntax highlights this analogy in that some of our constructions
%% are generalizations of set comprehension or power-sets and we use
%% similar notation.

While the restricted syntax developed in this paper does not express
some important concepts such as functor categories or opposite
categories, the restriction is natural in that it corresponds exactly to
virtual equipments, a well-understood notion of model that can express a
great deal of fundamental results and constructions in category
theory~\cite{riehl_verity_2022,Shulman13}.
%
Moreover, we can work around these unary/binary restrictions to some
extent by viewing the type theory as a domain-specific language embedded
in a metalanguage.  For example, while we cannot talk about functor
categories, we can state a theorem that quantifies over functors using
the meta-language's ``external'' universal quantifier (which does not
have automatic functoriality/naturality properties).  To support this,
\vett{} includes a third layer, an extensional dependent type theory in
the style of Martin-L\"of type theory. All of our ordered predicate
logic judgments are also indexed by a context from this dependent
type theory, and the type theory includes universe types for categories,
functors, profunctors and natural transformations.  This allow us to
formalize theorems the object logic is too restrictive to encode, analogous to
2-level~\cite{voevodsky13hts,altenkirch+16strict} or indexed type
theories~\cite{isaev21indexed,cervesatopfenning02llf,vakar15linear,krishnaswami+15linear}.

@@

While we emphasize the applications to enriched and internal category
theory in this work, there is potential for more direct application to
programming language semantics.
%
Ordinary predicate logic is the foundation for proof-theoretic presentations of
logical relations, such as Abadi-Plotkin logic for parametricity and
LSLR and Iris for step-indexed logical relations proofs
\cite{abadi-plotkin,lslr,iris}.
%
We conjecture that \vett{} might similarly serve as the foundation for
a logic of \emph{ordered} structures, which abound in applications:
rewriting and approximation relations can both be modeled as orderings
and logical relations involving these structures are proven to respect
orderings; operational logical relations must be downward-closed and
approximation relations should satisfy transitivity.
%
Just as LSLR and Iris release the user from the syntactic burden of
explicit step-indexing, \vett{} may be used to release the user from
the syntactic burden of proving downward-closure or transitivity
side-conditions.
%
Additionally, \vett{} may serve as the basis of a future domain specific
proof assistant for category theoretic proofs.  To pilot-test this, we
have formalized the syntax of \vett{} in Agda 2.6.2.2, using the rewrite
mechanism to make \vett's substitution and $\beta$-reduction rules
definitional
equalities.\footnote{\url{https://github.com/maxsnew/virtual-equipments/blob/master/agda/STC.agda}}
We have used this lightweight implementation to check a number of
examples.

\textbf{Basics of Profunctors.}  While we assume the reader has some
background knowledge of category theory, we briefly define profunctors,
which are not included in many introductory texts.  Recall that a
category $\cat C$ has a collection of objects and morphisms with
identity and composition, and a functor $F : \cat C \to \cat D$ is a
function on objects and a function on morphisms that preserves identity
and composition.  A category can be thought of as a generalization of a
preordered set, which has a set of elements and a binary \emph{relation}
on its objects satisfying reflexivity and transitivity.  A category is
then a \emph{proof-relevant preorder}, where morphisms are the proofs of
ordering, and the reflexivity and transitivity proofs must satisfy identity and
unit equations. A functor is then a \emph{proof-relevant monotone
  function}.  Given categories $\mathcal C$ and $\mathcal D$, a
profunctor $R$ from $\mathcal C$ to $\mathcal D$, written $R : \cat C
\pto \cat D$ is a functor $R : {\cat C}^{op} \times {\cat D} \to \Set$.
Because a profunctor outputs a $\Set$ rather than a proposition, it is
itself a \emph{proof-relevant relation}.  Thinking of categories as
proof-relevant preorders, functoriality says that the profunctor is
downward-closed in $\cat C$ and upward-closed in $\cat D$.  Given
profunctors $R, S : \cat C \pto \cat D$, a homomorphism from $R$ to $S$
is a natural transformation, which in the preordered setting is
simply an implication of relations.

@@

Finally we have several forms of context which are used in the theory.
The contexts $\Gamma$ of term variables with their types are as usual;
we write ${\Gamma \isadtctx}$ to indicate that a context is well-formed.
We name the remaining contexts after the judgements that they are used
by.  The set contexts $\Gamma \vdash \Xi \boundary$, which will be used
to type-check sets, contain object variables with their categories. The
two forms of set context are $\alpha : \cat C$, containing one variable
that can be used both contravariantly and covariantly, and $\alpha :
\cat C ; \beta : \cat D$, containing a contravariant variable $\alpha$
and covariant variable $\beta$.  Finally, the transformation contexts
$\Gamma \vdash \Phi \isavectx$ contain element variables with their sets, alternating with those
sets' object variables with their categories.  A typical $\Phi$ has the
shape
\[
\alpha_1 : \cat C_1, x_1:R_1(\alpha_1,\alpha_2), \alpha_2 : \cat C_2,
x_2:R_2(\alpha_2,\alpha_3), \ldots, R_n(\alpha_n,\alpha_{n+1}),
\alpha_{n+1} : \cat C_{n+1}
\]
and represents the composition of the ``relations'' $R_1, R_2, R_3, \ldots,
R_n$.  We write $d^-(\Phi)$ for the first category variable in $\Phi$
(which we regard as the negative or contravariant position), $d^+(\Phi)$
for the last category variable in $\Phi$ (which we regard as the
positive or covariant position), and $\Phi_1 \jnctx \Phi_2$ for the append
of two transformation contexts where the last variable in $\Phi_1$ is equal to
the first variable in $\Phi_2$.  Formal inductive definitions are in the
appendix, but intuitively: 
\[
\begin{array}{ccl}
d^-(\alpha_1 : \cat C_1, x_1:R_1(\alpha_1,\alpha_2), \ldots, x_n:R_n(\alpha_n,\alpha_n), \alpha_{n+1} : \cat C_{n+1}) & = & \alpha_1 : \cat C_1\\
d^+(\alpha_1 : \cat C_1, x_1:R_1(\alpha_1,\alpha_2), \ldots, x_n:R_n(\alpha_n,\alpha_n), \alpha_{n+1} : \cat C_{n+1}) & = & \alpha_{n+1}:\cat C_{n+1}\\
(\Phi_1,\beta:\cat D) \jnctx (\beta:\cat D, \Phi_2) & = & \Phi_1, \beta : \cat D, \Phi_2
\end{array}
\]

>>>>

\begin{document}

\newif\ifextended
\extendedfalse
\newcommand{\theappendix}{\ifextended the appendix \else the extended version \cite{extended-version}\fi}

%% \newtheorem{theorem}{Theorem}
%% \newtheorem{lemma}{Lemma}
\newtheorem{construction}{Construction}
%% \newtheorem{definition}{Definition}
\newcommand{\vett}{VETT}
\newcommand{\ohol}{\vett}
\newcommand{\fVDC}{\textrm{fVDCs}}
\newcommand{\VMJ}{\textrm{VM}_J}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\pto}{\nrightarrow}
\newcommand{\pfrom}{\nleftarrow}
\newcommand{\vcat}{\mathcal}
\newcommand{\cat}{\mathbbm}
\newcommand{\isaSet}{\,\,\textrm{Set}}
\newcommand{\isaTy}{\,\,\textrm{Type}}
\newcommand{\isaCat}{\,\,\textrm{Cat}}
\newcommand{\isSmall}{\,\,\textrm{Small}}
\newcommand{\Set}{\textrm{Set}}
\newcommand{\MonCat}{\textrm{MonCat}}
\newcommand{\SymMonCat}{\textrm{SymMonCat}}
\newcommand{\Syn}{\textrm{Syn}}
\newcommand{\vtkmnd}{\mathbb{K}\text{Mod} (\vcat{V},T)}
\newcommand{\rmod}{\text{RMod}}
\newcommand{\lmod}{\text{LMod}}

@@

\newcommand{\odotexists}[1]{\mathop{\overset{\exists #1}\odot}}
\newcommand{\tensorexistsXwithYandZ}[3]{#2 \odotexists{#1} #3}
\newcommand{\tensorintroofXandY}[2]{\textrm{pair}_\odot(#1,#2)}
\newcommand{\tensorintroatXwithYandZ}[3]{(#2,#1,#3)}
\newcommand{\tensorelimWkontZ}[2]{\textrm{ind}_{\odot}(#2;#1)}

@@

\title{A Formal Logic for Formal Category Theory}
\author{Max S. New\inst{1,2}\orcidID{0000-0001-8141-195X} and Daniel R. Licata\inst{2}}
\institute{University of Michigan \and Wesleyan University}

@@

\begin{abstract}
  We present a domain-specific type theory for constructions and proofs
  in category theory. The type theory axiomatizes notions of category,
  functor, profunctor and a generalized form of natural
  transformations. The type theory imposes an ordered linear restriction
  on standard predicate logic, which guarantees that all functions
  between categories are functorial, all relations are profunctorial,
  and all transformations are natural by construction, with no separate
  proofs necessary. Important category-theoretic proofs such as the
  Yoneda lemma and Co-yoneda lemma become simple type-theoretic proofs
  about the relationship between unit, tensor and (ordered) function
  types, and can be seen to be ordered refinements of theorems in
  predicate logic.  The type theory is sound and complete for a
  categorical model in \emph{virtual equipments}, which model both
  internal and enriched category theory. While the proofs in our type
  theory look like standard set-based arguments, the syntactic
  discipline ensure that all proofs and constructions carry over to
  enriched and internal settings as well.
\end{abstract}

@@

Category theory is a branch of mathematics that studies
higher-dimensional typed algebraic structures.
%
Originally developed for applications to homological algebra, it was
quickly discovered that categorical structures were common in logic and
computer science.  Formal systems like logics, type theories and
programming languages typically have sound and complete models given by
notions of structured categories~\cite{Lawvere69,LambekScott,moggi}.
%
This Curry-Howard-Lambek correspondence~
applies to simply typed lambda calculus~\cite{LambekScott},
computational lambda calculus~\cite{moggi}, linear logic~\cite{girard}
dependent type theory~\cite{cartmell1986,seely_1984}, and many other type theories
designed based on category-theoretic semantics.
%
The syntax of a type theory should present an initial object in its
category of models, a category-theoretic reformulation of logical
soundness and completeness.

While this research program has been quite successful,
category-theoretic notions can be overwhelming for beginners.
%
In a traditional set-theoretic formulation, notions such as adjoint
functors and limits produce a proliferation of ``naturality'' and
``functoriality'' side-conditions that must be discharged.
%
For example, when constructing an adjoint pair of
functors between two categories, a na\"ive approach would define all of
the data of the action on objects, action on arrows, prove the
functoriality of such actions, as well as construct two families of
transformations, prove they are natural and then finally proving a pair
of equalities relating compositions of natural transformations.
%
Carrying out these proofs explicitly is quite tedious and many
newcomers are left with the impression that category theory is full of
long, but ultimately trivial constructions.
%
This complexity is compounded when moving from ordinary category
theory to enriched and internal category theory, where constructions
must be additionally proven continuous, monotone, etc, in addition to
natural or functorial.
%
However, these generalizations are often exactly what is needed for programming
language applications; for example, domain-, metric- and
step-index-enriched categories have been used to model recursive
programming languages and internal categories have been used to model
parametricity and gradual
typing~\cite{wand197913,topos-of-trees,reflexive-graphs,double-cats-gradual-typing}.

Fortunately, the tools of category theory itself can be employed to
simplify this complexity, specifically the tools of \emph{higher}
category theory.
%
As an analogy in differential calculus, when an adept analyst writes
down a function, they do not expand out the $\epsilon\mathord{-}\delta$
definition of continuity for a function and proceed from first
principles, but rather use certain \emph{syntactic principles} for
defining functions that are continuous by construction --- e.g. that
composition of continuous functions is continuous.
%
Similar principles apply to category theory itself: functors and
natural transformations are closed under composition and whiskering
operations, and experienced category theorists rely on these syntactic
principles to eliminate the tedium of explicit proofs.
%
In the case of category theory, these principles can be formalized
using algebraic structures such as 2-categories, bicategories, Yoneda
structures, (virtual) double categories, pro-arrow
equipments~\cite{bicategories,proarrow-equipments,STREET1978350,LEINSTER2002391,Cruttwell2010},
an approach known as \emph{formal category theory}.
%
In these structures, rather than defining notions of category, functor
and natural transformation from first principles, they are axiomatized
in a manner similar to how a category axiomatizes a notion of space
and homomorphism.
%
Proofs in formal category theory apply to enriched and internal
settings, which are instances of the formal axioms.
%
A downside is that these algebraic structures are quite complicated,
and practitioners typically employ either an algebraic combinator
syntax (formalized in \citet{CURIEN1986188}) or a
2-dimensional diagrammatic language that can be quite beautiful and
elegant, but is also somewhat removed from the traditional formulation
of category theory in terms of sets and functions.

@@

%% The type-theoretic syntax allows us to write all of our 
%% constructions are written as if they were
%% ordinary functors and profunctors in set-based category theory.

@@

The syntax of \vett{} is an indexed, ordered linear, proof-relevant
variant of predicate logic over a unary type theory.
%
Just as a predicate logic has a notion of type, term, relation and
implication, \vett{} is based on four analogous category-theoretic
concepts: categories, functors, profunctors and natural transformations
of profunctors.
%
Categories are treated like types, and the unary functors we consider in
this paper are each represented by a term whose type is a category and
whose one free variable ranges over a category.
%
The analog of a relation is a \emph{profunctor} (defined below), which
is written like a set with free category variables.  Like the
restriction to unary functors, we restrict to profunctors with two free
variables.
%
The logic is proof-relevant in that the implications of relations are
generalized to natural transformations of profunctors, and we use a
$\lambda$-calculus notation to describe these ``proof terms''.
%
This analogy to predicate logic can be made formal: any construction in
\vett{} can be erased to a corresponding construction or proof in
predicate logic, as sets, functions, relations, and implication of
relations define a (somewhat degenerate) virtual equipment.
%% %
%% Our syntax highlights this analogy in that some of our constructions
%% are generalizations of set comprehension or power-sets and we use
%% similar notation.

While the restricted syntax developed in this paper does not express
some important concepts such as functor categories or opposite
categories, the restriction is natural in that it corresponds exactly to
virtual equipments, a well-understood notion of model that can express a
great deal of fundamental results and constructions in category
theory~\cite{riehl_verity_2022,Shulman13}.
%
Moreover, we can work around these unary/binary restrictions to some
extent by viewing the type theory as a domain-specific language embedded
in a metalanguage.  For example, while we cannot talk about functor
categories, we can state a theorem that quantifies over functors using
the meta-language's ``external'' universal quantifier (which does not
have automatic functoriality/naturality properties).  To support this,
\vett{} includes a third layer, an extensional dependent type theory in
the style of Martin-L\"of type theory. All of our ordered predicate
logic judgments are also indexed by a context from this dependent
type theory, and the type theory includes universe types for categories,
functors, profunctors and natural transformations.  This allow us to
formalize theorems the object logic is too restrictive to encode, analogous to
2-level~\cite{voevodsky13hts,altenkirch+16strict,PALMGREN2019102715} or indexed type
theories~\cite{isaev21indexed,cervesatopfenning02llf,vakar15linear,krishnaswami+15linear}.

@@

While we emphasize the applications to enriched and internal category
theory in this work, there is potential for more direct application to
programming language semantics.
%
Ordinary predicate logic is the foundation for proof-theoretic presentations of
logical relations, such as Abadi-Plotkin logic for parametricity and
LSLR and Iris for step-indexed logical relations proofs
\cite{abadi-plotkin,lslr,iris}.
%
We conjecture that \vett{} might similarly serve as the foundation for
a logic of \emph{ordered} structures, which abound in applications:
rewriting and approximation relations can both be modeled as orderings
and logical relations involving these structures are proven to respect
orderings; operational logical relations must be downward-closed and
approximation relations should satisfy transitivity.
%
Just as LSLR and Iris release the user from the syntactic burden of
explicit step-indexing, \vett{} may be used to release the user from
the syntactic burden of proving downward-closure or transitivity
side-conditions.
%
Additionally, \vett{} may serve as the basis of a future domain specific
proof assistant for category-theoretic proofs.  To pilot-test this, we
have formalized the syntax of \vett{} in Agda 2.6.2.2, using the rewrite
mechanism to make \vett's substitution and $\beta$-reduction rules
definitional
equalities.\footnote{\url{https://github.com/maxsnew/virtual-equipments/blob/master/agda/STC.agda}}
We have used this lightweight implementation to check a number of
examples.

\textbf{Basics of Profunctors.}  While we assume the reader has some
background knowledge of category theory, we briefly define profunctors,
which are not included in many introductory texts.  Recall that a
category $\cat C$ has a collection of objects and morphisms with
identity and composition, and a functor $F : \cat C \to \cat D$ is a
function on objects and a function on morphisms that preserves identity
and composition.  A category can be thought of as a generalization of a
preordered set, which has a set of elements and a binary \emph{relation}
on its objects satisfying reflexivity and transitivity.  A category is
then a \emph{proof-relevant preorder}, where morphisms are the proofs of
ordering, and the reflexivity and transitivity proofs must satisfy identity and
unit equations. A functor is then a \emph{proof-relevant monotone
  function}.  Given categories $\mathcal C$ and $\mathcal D$, a
profunctor $R$ from $\mathcal C$ to $\mathcal D$, written $R : \cat C
\pto \cat D$ is a functor $R : {\cat C}^{o} \times {\cat D} \to \Set$\footnote{${\cat C}^o$ is the notation we use for the opposite category of $\cat C$}.
Because a profunctor outputs a $\Set$ rather than a proposition, it is
itself a \emph{proof-relevant relation}.  Thinking of categories as
proof-relevant preorders, functoriality says that the profunctor is
downward-closed in $\cat C$ and upward-closed in $\cat D$.  Given
profunctors $R, S : \cat C \pto \cat D$, a homomorphism from $R$ to $S$
is a natural transformation, which in the preordered setting is
simply an implication of relations.

@@

Finally we have several forms of context which are used in the theory.
The contexts $\Gamma$ of term variables with their types are as usual;
we write ``${\Gamma \isadtctx}$'' to indicate that a context is well-formed.
We name the remaining contexts after the judgements that they are used
by.  The set contexts $\Xi$, which will be used
to type-check sets, contain object variables with their categories. The
two forms of set context are $\alpha : \cat C$, containing one variable
that can be used both contravariantly and covariantly, and $\alpha :
\cat C ; \beta : \cat D$, containing a contravariant variable $\alpha$
and covariant variable $\beta$.  Finally, the transformation contexts
$\Phi$ contain element variables with their sets, alternating with those
sets' object variables with their categories.  A typical $\Phi$ has the
shape
\[
\alpha_1 : \cat C_1, x_1:R_1(\alpha_1,\alpha_2), \alpha_2 : \cat C_2,
x_2:R_2(\alpha_2,\alpha_3), \ldots, R_n(\alpha_n,\alpha_{n+1}),
\alpha_{n+1} : \cat C_{n+1}
\]
and represents the composition of the ``relations'' $R_1, R_2, R_3,
\ldots, R_n$.  We write $d^-(\Phi)$ for the first category variable in
$\Phi$ (which we regard as the negative or contravariant position),
$d^+(\Phi)$ for the last category variable in $\Phi$ (which we regard
as the positive or covariant position) and use the notation $d^\pm\Xi$
with the same meaning. We write $\Phi_1 \jnctx \Phi_2$ for the append
of two transformation contexts, which is only well-formed when the
last variable in $\Phi_1$ is equal to the first variable in
$\Phi_2$. Formal inductive definitions are in the appendix, but
intuitively:
\[
\begin{array}{ccl}
d^-(\alpha_1 : \cat C_1, x_1:R_1(\alpha_1,\alpha_2), \ldots, x_n:R_n(\alpha_n,\alpha_n), \alpha_{n+1} : \cat C_{n+1}) & = & \alpha_1 : \cat C_1\\
d^+(\alpha_1 : \cat C_1, x_1:R_1(\alpha_1,\alpha_2), \ldots, x_n:R_n(\alpha_n,\alpha_n), \alpha_{n+1} : \cat C_{n+1}) & = & \alpha_{n+1}:\cat C_{n+1}\\
(\Phi_1,\beta:\cat D) \jnctx (\beta:\cat D, \Phi_2) & = & \Phi_1, \beta : \cat D, \Phi_2
\end{array}
\]

%% %
%% First, we have the types and terms of the logic over our category
%% theory. Types include the usual dependent sum, product and identity
%% types of dependent type theory, as well as several forms that allow us
%% to abstract over the category-theoretic constructions: a type $\Cat$
%% of categories, a type $\varr{\cat C}{\cat D}$ for functors from $\cat
%% C$ to $\cat D$, a type $\harr{\cat C}{\cat D}$ for profunctors
%% contravariant in $\cat C$ and covariant in $\cat D$, and finally a
%% type $\pendallXdotY {\alpha:{\cat C}} R$ that, when combined with
%% other constructions on profunctors, allows for abstraction over
%% arbitrary natural transformations.
%% %
%% Terms $L,M,N$ describe the elements of our logic's types, and include
%% the usual constructions as well as forms for defining categories,
%% functors, profunctors and natural transformations.
%% %
%% Next, we have a syntax for categories, which are written using a
%% blackboard font $\cat C, \cat D, \cat E$. Categories include terms $M$
%% of the category type (allowing for abstraction), in addition to
%% primitives for product and unit categories.
%% %
%% Next, we have a syntax for objects of a category, which, because they
%% are typed with respect to an input variable, provide a syntax for
%% functors.
%% %
%% The objects we can define are object variables, application of a
%% functor $M a$, and pairs of objects (for the product category) and the
%% unique object of the unit category.
%% %
%% Next, $P,Q,R$ stand for \emph{sets}, but similar to objects, they are
%% always typed with respect to a contravariant and covariant variable,
%% and so denote profunctors.
%% %
%% These include first the multiplicatives: the hom set
%% $\punitinXfromYtoZ {\cat C} a b$ representing the morphisms from $a$
%% to $b$ in $\cat C$, a tensor product $\tensorexistsXwithYandZ
%% {\beta:\cat D} P Q$, and two kinds of ``ordered function types''
%% $\homrallXYtoZ \beta P Q$ and $\homlallXYtoZ \alpha R S$. 
%% %
%% We also include application of a profunctor term $M$ to two objects,
%% and cartesian product and unit sets.
%% %
%% Next, we have transformations $s,t,u$, which are typed with respect
%% to (0 or more) input and (one) output set.
